---
title: "perigrine_distribution"
author: "Shale"
date: "1/5/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Ben
https://ucsb.zoom.us/j/3133990354
Sam
https://ucsb.zoom.us/j/4565706927

```{r}
library(librarian)

librarian::shelf(dismo, dplyr, DT, ggplot2, here, 
                 htmltools, leaflet, mapview, purrr, 
                 raster, readr, rgbif, rgdal, rJava, 
                 sdmpredictors, sf, spocc, tidyr)

select <- dplyr::select # overwrite raster::select
options(readr.show_col_types = FALSE)
```


```{r}
set.seed(15)

# directory to store data
dir_data <- here("data")
dir.create(dir_data, showWarnings = F)
```

## Species Data

```{r}
obs_csv <- file.path(dir_data, "obs.csv")
obs_geo <- file.path(dir_data, "obs.geojson")
redo <- TRUE
usa_hull =st_as_sf(rgeos::readWKT("POLYGON ((-125.33203125 39.232253141714885, -114.60937499999999 25.799891182088334, -79.453125 24.686952411999155, -79.453125 31.052933985705163, -57.65624999999999 47.27922900257082, -93.8671875 49.724479188712984, -126.73828125 49.61070993807422, -125.33203125 39.232253141714885))"))

if (!file.exists(obs_geo) | redo){
  # get species occurrence data from GBIF with coordinates
  (res <- spocc::occ(
    query = 'Falco peregrinus', 
    from = 'gbif', has_coords = T,
    geometry = usa_hull,
    limit = 20000))
  
  # extract data frame from result
  df <- res$gbif$data[[1]] 
  readr::write_csv(df, obs_csv)
  
  # convert to points of observation from lon/lat columns in data frame
  obs <- df %>% 
    sf::st_as_sf(
      coords = c("longitude", "latitude"),
      crs = st_crs(4326)) %>% 
    select(prov, key) # save space (joinable from obs_csv)
  sf::write_sf(obs, obs_geo, delete_dsn=T)
}
obs <- sf::read_sf(obs_geo)
nrow(obs) # number of rows
```

```{r}
# show points on map
mapview::mapview(obs, map.types = "OpenTopoMap")
```

Though I set the limit to 20,000 observations in the US (and there are clearly more than this, as the limit was reached), there are 1,738,469 global occurrences of _Falco peregrinus_ on gbif.org as of 2022/01/05.

There are several sightings over the open ocean, which are probably erroneous. Some may not be (given that there are many sightings on nearby islands, and it is possible some birds may have been observed in flight between landmasses, particularly juveniles in search of new territory [`3399467453` in particular fits this possibility]) but to be safe I will remove all sightings not visibly over land. To remove these suspected false sightings:

```{r}
obs = obs %>% subset(!key %in% c(3407992796, 3399467453, 3415497266, 
                                 3394760531, 3332604472))
# Most of these fall outside of the hull used above, which is why only 1 observation was removed:
nrow(obs)
```

## Environmental Data

```{r}
dir_env <- file.path(dir_data, "env")

# set a default data directory
options(sdmpredictors_datadir = dir_env)

# choosing terrestrial
env_datasets <- sdmpredictors::list_datasets(terrestrial = TRUE, marine = FALSE)

# show table of datasets
env_datasets %>% 
  select(dataset_code, description, citation)

# choose datasets for a vector
env_datasets_vec <- c("WorldClim", "ENVIREM")

# get layers
env_layers <- sdmpredictors::list_layers(env_datasets_vec)
```

```{r}
mapview(
  list(obs, usa_hull))
```

```{bash, eval=FALSE}
for item in WC_alt WC_bio1 WC_bio2 ER_tri ER_topoWet
do
	curl -v -k https://www.lifewatch.be/sdmpredictors/${item}.tif --output data/${item}.tif
done
```

```{r}
env_layer_file_vec <- paste0("data/", list.files(here("data"), pattern = "*.tif"), sep = "")

env_stack <- raster::stack(env_layer_file_vec)

plot(env_stack, nc=2)
```

For some reason this is yielding blank plots:

```{r}
env_stack_grd <- file.path(dir_data, "env_stack.grd")

# this if was throwing errors, so i just removed it
# if (!file.exists(env_stack_grd) | redo){
  obs_hull_sp <- sf::as_Spatial(usa_hull)
  env_stack <- raster::mask(env_stack, obs_hull_sp) %>% 
    raster::crop(extent(obs_hull_sp))
  writeRaster(env_stack, env_stack_grd, overwrite=T)  
# }
env_stack <- stack(env_stack_grd)

# show map
# mapview(obs) + 
#   mapview(env_stack, hide = T) # makes html too big for Github
plot(env_stack, nc=2)
```

